---
title: "Poster Outputs: Tree-based Models and Cluster Analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Packages

```{r}
library(tidyverse)
library(randomForest)
library(rpart)
library(rpart.plot)
library(cluster)

set.seed(43)

if (!dir.exists("poster_outputs")) {
  dir.create("poster_outputs")
}
```

# 2. Load and prepare data

```{r}
data <- read.csv("student_productivity.csv")

data <- data %>%
  drop_na() %>%
  mutate(gender = as.factor(gender))

str(data)
summary(data$productivity_score)
```

# 3. Train-test split

```{r}
set.seed(43)
train_index <- sample(1:nrow(data), size = 0.8 * nrow(data))

train <- data[train_index, ]
test  <- data[-train_index, ]

nrow(train)
nrow(test)
```

# 4. Tree-based regression models

The target variable is `productivity_score`, so this is a regression task.
`student_id` is removed because it is only an identifier.
`final_grade` is also removed because it may be too close to the outcome and could weaken the interpretation of behavioural predictors.

```{r}
model_formula <- productivity_score ~
  age +
  gender +
  study_hours_per_day +
  sleep_hours +
  phone_usage_hours +
  social_media_hours +
  youtube_hours +
  gaming_hours +
  breaks_per_day +
  coffee_intake_mg +
  exercise_minutes +
  assignments_completed +
  attendance_percentage +
  stress_level +
  focus_score
```

## 4.1 Regression tree

```{r}
set.seed(43)
reg_tree <- rpart(
  model_formula,
  data = train,
  method = "anova"
)

reg_tree
```

```{r, fig.width=10, fig.height=7}
rpart.plot(
  reg_tree,
  main = "Regression Tree for Student Productivity",
  type = 2,
  extra = 101,
  fallen.leaves = TRUE,
  cex = 0.7
)
```

```{r}
png("poster_outputs/regression_tree.png", width = 1600, height = 1100, res = 150)
rpart.plot(
  reg_tree,
  main = "Regression Tree for Student Productivity",
  type = 2,
  extra = 101,
  fallen.leaves = TRUE,
  cex = 0.7
)
dev.off()
```

## 4.2 Regression tree performance

```{r}
tree_pred <- predict(reg_tree, newdata = test)

tree_mse <- mean((test$productivity_score - tree_pred)^2)
tree_rmse <- sqrt(tree_mse)
tree_r2 <- cor(test$productivity_score, tree_pred)^2

tree_results <- data.frame(
  Model = "Regression Tree",
  Test_MSE = tree_mse,
  Test_RMSE = tree_rmse,
  Test_R2 = tree_r2
)

tree_results
```

## 4.3 Random forest regression

```{r}
set.seed(43)
rf_model <- randomForest(
  model_formula,
  data = train,
  mtry = 3,
  ntree = 500,
  importance = TRUE
)

rf_model
```

```{r}
rf_pred <- predict(rf_model, newdata = test)

rf_mse <- mean((test$productivity_score - rf_pred)^2)
rf_rmse <- sqrt(rf_mse)
rf_r2 <- cor(test$productivity_score, rf_pred)^2

rf_results <- data.frame(
  Model = "Random Forest, mtry = 3",
  Test_MSE = rf_mse,
  Test_RMSE = rf_rmse,
  Test_R2 = rf_r2
)

rf_results
```

## 4.4 Tune mtry for random forest

```{r}
set.seed(43)

mtry_values <- 2:10
mtry_results <- data.frame(
  mtry = mtry_values,
  OOB_RMSE = NA
)

for (i in seq_along(mtry_values)) {
  temp_rf <- randomForest(
    model_formula,
    data = train,
    mtry = mtry_values[i],
    ntree = 300,
    importance = TRUE
  )
  mtry_results$OOB_RMSE[i] <- sqrt(tail(temp_rf$mse, 1))
}

mtry_results
best_mtry <- mtry_results$mtry[which.min(mtry_results$OOB_RMSE)]
best_mtry
```

```{r, fig.width=8, fig.height=5}
mtry_plot <- ggplot(mtry_results, aes(x = mtry, y = OOB_RMSE)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "Random Forest Tuning",
    subtitle = "Choosing mtry using out-of-bag RMSE",
    x = "mtry: number of predictors randomly considered at each split",
    y = "OOB RMSE"
  ) +
  theme_minimal(base_size = 13)

mtry_plot

ggsave(
  filename = "poster_outputs/random_forest_mtry_tuning.png",
  plot = mtry_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```

## 4.5 Final tuned random forest

```{r}
set.seed(43)
rf_final <- randomForest(
  model_formula,
  data = train,
  mtry = best_mtry,
  ntree = 500,
  importance = TRUE
)

rf_final
```

```{r}
rf_final_pred <- predict(rf_final, newdata = test)

rf_final_mse <- mean((test$productivity_score - rf_final_pred)^2)
rf_final_rmse <- sqrt(rf_final_mse)
rf_final_r2 <- cor(test$productivity_score, rf_final_pred)^2

rf_final_results <- data.frame(
  Model = paste0("Tuned Random Forest, mtry = ", best_mtry),
  Test_MSE = rf_final_mse,
  Test_RMSE = rf_final_rmse,
  Test_R2 = rf_final_r2
)

rf_final_results
```

## 4.6 Model comparison for poster

```{r}
model_comparison <- bind_rows(
  tree_results,
  rf_results,
  rf_final_results
)

model_comparison

write.csv(model_comparison, "poster_outputs/model_comparison.csv", row.names = FALSE)
```

```{r, fig.width=8, fig.height=5}
model_plot <- ggplot(model_comparison, aes(x = reorder(Model, Test_RMSE), y = Test_RMSE)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Tree-based Model Comparison",
    subtitle = "Lower test RMSE indicates better prediction accuracy",
    x = "Model",
    y = "Test RMSE"
  ) +
  theme_minimal(base_size = 13)

model_plot

ggsave(
  filename = "poster_outputs/tree_model_comparison_rmse.png",
  plot = model_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```

## 4.7 Actual vs predicted plot

```{r, fig.width=7, fig.height=6}
actual_pred_df <- data.frame(
  Actual = test$productivity_score,
  Predicted = rf_final_pred
)

actual_pred_plot <- ggplot(actual_pred_df, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.35) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Random Forest: Actual vs Predicted Productivity Score",
    subtitle = "Points closer to the dashed line indicate better predictions",
    x = "Actual productivity score",
    y = "Predicted productivity score"
  ) +
  theme_minimal(base_size = 13)

actual_pred_plot

ggsave(
  filename = "poster_outputs/random_forest_actual_vs_predicted.png",
  plot = actual_pred_plot,
  width = 7,
  height = 6,
  dpi = 300
)
```

## 4.8 Random forest variable importance

```{r}
importance_matrix <- importance(rf_final)

importance_df <- data.frame(
  Variable = rownames(importance_matrix),
  IncMSE = importance_matrix[, "%IncMSE"],
  IncNodePurity = importance_matrix[, "IncNodePurity"],
  row.names = NULL
) %>%
  arrange(desc(IncMSE))

importance_df

write.csv(importance_df, "poster_outputs/random_forest_variable_importance.csv", row.names = FALSE)
```

```{r, fig.width=9, fig.height=6}
importance_plot <- importance_df %>%
  slice_max(order_by = IncMSE, n = 10) %>%
  ggplot(aes(x = reorder(Variable, IncMSE), y = IncMSE)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Predictors of Student Productivity",
    subtitle = "Random forest variable importance based on % increase in MSE",
    x = "Predictor",
    y = "% Increase in MSE"
  ) +
  theme_minimal(base_size = 13)

importance_plot

ggsave(
  filename = "poster_outputs/random_forest_variable_importance.png",
  plot = importance_plot,
  width = 9,
  height = 6,
  dpi = 300
)
```

# 5. Cluster analysis

Clustering is unsupervised learning. Here, the goal is not to predict `productivity_score`, but to explore whether students can be grouped into behavioural profiles based on study habits, digital usage, wellbeing, and academic engagement.

For clustering, only numeric behavioural features are used. The target variable `productivity_score`, `final_grade`, and identifier `student_id` are removed.

```{r}
cluster_data <- data %>%
  select(
    age,
    study_hours_per_day,
    sleep_hours,
    phone_usage_hours,
    social_media_hours,
    youtube_hours,
    gaming_hours,
    breaks_per_day,
    coffee_intake_mg,
    exercise_minutes,
    assignments_completed,
    attendance_percentage,
    stress_level,
    focus_score
  )

cluster_scaled <- scale(cluster_data)

summary(cluster_data)
```

## 5.1 PCA for visualising clusters

```{r}
pca_model <- prcomp(cluster_scaled, center = FALSE, scale. = FALSE)

pca_df <- data.frame(
  PC1 = pca_model$x[, 1],
  PC2 = pca_model$x[, 2],
  productivity_score = data$productivity_score
)

pca_var <- round(100 * summary(pca_model)$importance[2, 1:2], 1)
pca_var
```

```{r, fig.width=7, fig.height=6}
pca_base_plot <- ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.25) +
  labs(
    title = "Students in PCA Space Before Clustering",
    x = paste0("PC1 (", pca_var[1], "% variance)"),
    y = paste0("PC2 (", pca_var[2], "% variance)")
  ) +
  theme_minimal(base_size = 13)

pca_base_plot

ggsave(
  filename = "poster_outputs/pca_before_clustering.png",
  plot = pca_base_plot,
  width = 7,
  height = 6,
  dpi = 300
)
```

## 5.2 K-means clustering: choose k using within-cluster criterion

```{r}
set.seed(43)

k_values <- 2:8
kmeans_criterion <- data.frame(
  k = k_values,
  criterion = NA
)

for (i in seq_along(k_values)) {
  km <- kmeans(cluster_scaled, centers = k_values[i], nstart = 20)
  kmeans_criterion$criterion[i] <- 1 - km$tot.withinss / km$totss
}

kmeans_criterion
```

```{r, fig.width=8, fig.height=5}
kmeans_criterion_plot <- ggplot(kmeans_criterion, aes(x = k, y = criterion)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "K-means Criterion vs Number of Clusters",
    subtitle = "Higher values indicate more between-cluster separation",
    x = "Number of clusters, k",
    y = "1 - total within SS / total SS"
  ) +
  theme_minimal(base_size = 13)

kmeans_criterion_plot

ggsave(
  filename = "poster_outputs/kmeans_criterion_vs_k.png",
  plot = kmeans_criterion_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```

## 5.3 K-means clustering: choose k using silhouette coefficient

For computational stability, the silhouette coefficient is calculated on a random sample of 2,000 observations.

```{r}
set.seed(43)
sil_sample_size <- min(2000, nrow(cluster_scaled))
sil_index <- sample(1:nrow(cluster_scaled), sil_sample_size)
cluster_scaled_sample <- cluster_scaled[sil_index, ]
dist_sample <- dist(cluster_scaled_sample)

silhouette_results <- data.frame(
  k = k_values,
  avg_silhouette = NA
)

for (i in seq_along(k_values)) {
  km_sample <- kmeans(cluster_scaled_sample, centers = k_values[i], nstart = 20)
  sil <- silhouette(km_sample$cluster, dist_sample)
  silhouette_results$avg_silhouette[i] <- summary(sil)$avg.width
}

silhouette_results

best_k <- silhouette_results$k[which.max(silhouette_results$avg_silhouette)]
best_k
```

```{r, fig.width=8, fig.height=5}
silhouette_plot <- ggplot(silhouette_results, aes(x = k, y = avg_silhouette)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "Silhouette Coefficient vs Number of Clusters",
    subtitle = paste0("Best k based on sampled silhouette = ", best_k),
    x = "Number of clusters, k",
    y = "Average silhouette coefficient"
  ) +
  theme_minimal(base_size = 13)

silhouette_plot

ggsave(
  filename = "poster_outputs/silhouette_vs_k.png",
  plot = silhouette_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```

## 5.4 Final k-means model

```{r}
set.seed(43)
kmeans_final <- kmeans(cluster_scaled, centers = best_k, nstart = 20)

data_clustered <- data %>%
  mutate(cluster = factor(kmeans_final$cluster))

pca_cluster_df <- pca_df %>%
  mutate(cluster = factor(kmeans_final$cluster))

kmeans_final$size
kmeans_final$tot.withinss
```

```{r, fig.width=7, fig.height=6}
kmeans_pca_plot <- ggplot(pca_cluster_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.35) +
  labs(
    title = "K-means Student Behaviour Clusters",
    subtitle = paste0("Visualised using PCA; k = ", best_k),
    x = paste0("PC1 (", pca_var[1], "% variance)"),
    y = paste0("PC2 (", pca_var[2], "% variance)"),
    color = "Cluster"
  ) +
  theme_minimal(base_size = 13)

kmeans_pca_plot

ggsave(
  filename = "poster_outputs/kmeans_clusters_pca.png",
  plot = kmeans_pca_plot,
  width = 7,
  height = 6,
  dpi = 300
)
```

## 5.5 Cluster profiles

```{r}
cluster_profiles <- data_clustered %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    avg_productivity = mean(productivity_score),
    avg_study_hours = mean(study_hours_per_day),
    avg_sleep_hours = mean(sleep_hours),
    avg_phone_usage = mean(phone_usage_hours),
    avg_social_media = mean(social_media_hours),
    avg_youtube = mean(youtube_hours),
    avg_gaming = mean(gaming_hours),
    avg_breaks = mean(breaks_per_day),
    avg_coffee = mean(coffee_intake_mg),
    avg_exercise = mean(exercise_minutes),
    avg_assignments = mean(assignments_completed),
    avg_attendance = mean(attendance_percentage),
    avg_stress = mean(stress_level),
    avg_focus = mean(focus_score),
    .groups = "drop"
  )

cluster_profiles

write.csv(cluster_profiles, "poster_outputs/cluster_profiles.csv", row.names = FALSE)
```

```{r, fig.width=8, fig.height=5}
cluster_productivity_plot <- ggplot(data_clustered, aes(x = cluster, y = productivity_score)) +
  geom_boxplot() +
  labs(
    title = "Productivity Score by Behaviour Cluster",
    subtitle = "This compares productivity distributions across unsupervised clusters",
    x = "Cluster",
    y = "Productivity score"
  ) +
  theme_minimal(base_size = 13)

cluster_productivity_plot

ggsave(
  filename = "poster_outputs/productivity_by_cluster.png",
  plot = cluster_productivity_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```

```{r, fig.width=9, fig.height=6}
cluster_profile_long <- cluster_profiles %>%
  select(
    cluster,
    avg_study_hours,
    avg_sleep_hours,
    avg_phone_usage,
    avg_social_media,
    avg_gaming,
    avg_exercise,
    avg_attendance,
    avg_stress,
    avg_focus,
    avg_productivity
  ) %>%
  pivot_longer(
    cols = -cluster,
    names_to = "Feature",
    values_to = "Value"
  )

cluster_profile_plot <- ggplot(cluster_profile_long, aes(x = Feature, y = Value, fill = cluster)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Cluster Profiles",
    subtitle = "Average feature values in each student behaviour cluster",
    x = "Feature",
    y = "Average value",
    fill = "Cluster"
  ) +
  theme_minimal(base_size = 12)

cluster_profile_plot

ggsave(
  filename = "poster_outputs/cluster_profiles.png",
  plot = cluster_profile_plot,
  width = 9,
  height = 6,
  dpi = 300
)
```

# 6. Hierarchical clustering on a sample

Hierarchical clustering requires a distance matrix, which can become very large for 20,000 observations. Therefore, this section uses a random sample for visualisation.

```{r}
set.seed(43)
hc_sample_size <- min(500, nrow(cluster_scaled))
hc_index <- sample(1:nrow(cluster_scaled), hc_sample_size)
hc_data <- cluster_scaled[hc_index, ]
hc_dist <- dist(hc_data)

hc_complete <- hclust(hc_dist, method = "complete")
```

```{r, fig.width=10, fig.height=6}
plot(hc_complete, main = "Hierarchical Clustering: Complete Linkage", xlab = "", sub = "", cex = 0.4)
rect.hclust(hc_complete, k = best_k, border = "red")
```

```{r}
png("poster_outputs/hierarchical_clustering_dendrogram.png", width = 1600, height = 1000, res = 150)
plot(hc_complete, main = "Hierarchical Clustering: Complete Linkage", xlab = "", sub = "", cex = 0.4)
rect.hclust(hc_complete, k = best_k, border = "red")
dev.off()
```

# 7. Short poster-ready summaries

```{r}
cat("Tree-based model summary:\n")
cat("The tuned random forest achieved a test RMSE of", round(rf_final_rmse, 3),
    "and test R-squared of", round(rf_final_r2, 3), ".\n")
cat("The most important predictor was", importance_df$Variable[1], ".\n\n")

cat("Cluster analysis summary:\n")
cat("K-means clustering was used to identify student behaviour profiles.\n")
cat("The silhouette analysis selected k =", best_k, "clusters.\n")
cat("Cluster profiles were compared using average productivity, study hours, sleep, phone use, stress, and focus.\n")
```
